---
title: "Assignment 2"
author: "Alexia Salomons, Nathan Maxwell Jones, Yauheniya Makarevich, group 71"
date: "15 March 2023"
output: pdf_document
fontsize: 11pt
highlight: tango
---

```{r setup, include = FALSE}
# set up global R options
options(digits = 3)

# set up knitr global chunk options
knitr::opts_chunk$set(fig.height = 3)
```


## Exercise 1

```{r}
tree_df <- as.data.frame(read.table("data/treeVolume.txt", header=TRUE))
head(tree_df)
```

**a)** 
1-way ANOVA
```{r}
tree_df$type <- as.factor(tree_df$type)
tree_type_lm <- lm(volume~type, data=tree_df)
anova(tree_type_lm)
```
```{r}
summary(tree_type_lm)
```

Therefore tree type does not influence volume significantly. Because the factor has two levels, we can apply a t-test.
```{r}
mask <- tree_df$type == "beech"
t.test(tree_df$volume[mask], tree_df$volume[!mask])
```
This supports the result from the ANOVA test. The estimated volume for Beech trees is 30.2 and 35.2 for Oak trees.


**b)** 

```{r}
tree_type_d_lm <- lm(volume~type*diameter, data=tree_df)
anova(tree_type_d_lm)
```
The influence of diameter on volume is similar for both tree types since the interaction between type and diameter is not significant.


```{r}
tree_type_h_lm <- lm(volume~type*height, data=tree_df)
anova(tree_type_h_lm)
```
The influence of height on volume is different for both tree types since the interaction between type and diameter is significant. We can see that type alone does not have a signficant effect, but it does interact significantly with height.

**c)** 

```{r}
tree_type_all_lm <- lm(volume~type*diameter+height, data=tree_df)
anova(tree_type_all_lm)
```

**d)** 

```{r}

```

## Exercise 2

```{r, include=FALSE}
# Read data
```

**a)** 

```{r}

```

**b)** 

```{r}

```

**c)** 

```{r}

```

**d)** 

```{r}

```

## Exercise 3

```{r, include=FALSE}
# Read data
```

**a)** 

```{r}

```

**b)** 

```{r}

```

**c)** 

```{r}

```

**d)** 

```{r}

```

**e)** 

```{r}

```

## Exercise 4

```{r, include=FALSE}
# Read data
```

**a)** 

```{r}

```

**b)** 

```{r}

```

**c)** 

```{r}

```

