---
title: "Assignment 1"
author: "Alexia Salomons, Nathan Maxwell Jones, Yauheniya Makarevich, group 71"
date: "27 February 2023"
output: pdf_document
fontsize: 11pt
highlight: tango
---

## Exercise 1. 
The data set birthweight.txt contains the birthweights (in grams) of 188 newborn babies. Denote the underlying mean birthweight by $\mu$.

```{r}
birthweight <- readLines("data/birthweight.txt")
birthweight <- as.double(birthweight[2:length(birthweight)])
birthweight
```
```{r}
length(birthweight)
```

```{r}
birthweight_mean <- mean(birthweight)
birthweight_mean
```

**a)** Check normality of the data. Assuming normality (irrespective of your conclusion about normality), construct a bounded 96%-CI for $\mu$. Evaluate the sample size needed to provide that the length of the 96%-CI is at most 100. Compute a bootstrap 96%-CI for $\mu$ and compare it to the above CI.


```{r}
qqnorm(y = birthweight)
```

```{r}
hist(birthweight)
# add density line
# lines(density(birthweight), col="blue",lwd=2)
```

```{r}
shapiro.test(birthweight)
# H0 - normal distribution, H1 - not normal
```
NORMAL DISTRIBUTION!

Let's go for CI-96%:

```{r}
t.test(birthweight, conf.level = 0.96)
```
```{r}
B <- 1000
alpha <- 0.04
T_star <- numeric(B)

for(i in 1:B) {
  X_star <- sample(birthweight, replace = TRUE)
  T_star[i] <- mean(X_star)
}

T_star_q2 <- quantile(T_star, alpha/2)
T_star_q98 <- quantile(T_star, 1 - alpha/2)

c(2*birthweight_mean - T_star_q98, 2*birthweight_mean - T_star_q2)
```
```{r}
sum(T_star<T_star_q2)
```

<!-- #TODO: FIGURE OUT SAMPLE SIZE FOR HAVING CI INTERVAL LENGTH + 100 -->
<!-- 828 babies to get the CI -->

**b)**  An expert claims that the mean birthweight is bigger than 2800 gram. Verify this claim by using a relevant t-test, explain the meaning of the CI in the R-output for this test. Also propose and perform a suitable sign tests for this problem.

```{r}
t.test(birthweight, alternative = "greater", mu=2800)
```
We reject H0(p=0.01337), so H1 is true and mean of the sample is bigger than 2800. 
CI is infinite on right side, since the test is one-sided.


Binom test

H0: mean <= 2800,
H1: mean > 2800.

```{r}
greater_weight <- as.integer(birthweight > 2800)
binom.test(sum(greater_weight), length(greater_weight), p=0.5, alt="g")
```
We reject H0(p=0.03868), so H1 is true and mean of the sample is bigger than 2800. 

Both test confirmed the hypothesis that mean of the sample is bigger than 2800.

**c)** Propose a way to compute the powers of the t-test and sing test from b) at some $\mu$ > 2800, comment.
```{r}


```

**d)**  Let $p$ be the probability that birthweight of a newborn baby is less than 2600 gram. Using asymptotic normality, the expert computed the left end $\hat{p}=0.25$ of the confidence interval $[\hat{p_l}, \hat{p_r}]$ for $p$. Recover the whole confidence interval and its confidence level.
```{r}

```

**e)** The expert also reports that there were 34 male and 28 female babies among 62 who weighted less than 2600 gram, and 61 male and 65 female babies among the remaining 126 babies. The expert claims that the mean weight is different for male and female babies. Verify this claim by an appropriate test.


success: w > 2600
```{r}
prop.test(c(61, 65), c(95, 93))
```
We accept H0: p1-p2=0, where p1, p2 - proportions of the success in population.

## Exercise 2
A study tested whether cholesterol was reduced after using a certain brand of margarine as part of a low fat low cholesterol diet. The data set cholesterol.txt contains information on 18 people using margarine to reduce cholesterol: columns Before and After8weeks contain the cholesterol level (mmol/L) respectively before the diet and after 8 weeks on the diet.

```{r}
df <- as.data.frame(read.table("data/cholesterol.txt", header=TRUE))
head(df)
```


**a)** Make some relevant plots of this data set, comment on normality. Are there any inconsistencies in the data? Investigate whether the columns Before and After8weeks are correlated.
```{r}
diffs <- df[, 1] - df[, 2]
diffs
qqnorm(diffs)
```

```{r}
hist(diffs)
```
```{r}
shapiro.test(diffs)
```
Differences are normally distributed.

```{r}
shapiro.test(df[, 1])
shapiro.test(df[, 2])
```
```{r}
cor.test(df[, 1], df[, 2], method="pearson")
```




## Exercise 1 
Below is a template for reporting the exercises from the assignments.

**a)** Here are some consequitive R-commands.
```{r}
x=rep(c("A","B"),each=5); x
sample(x)
x=rnorm(100)
```
Now the same code chunk but with all the output collapsed into signle block.
```{r, collapse=TRUE}
x=rep(c("A","B"),each=5); x
sample(x)
x=rnorm(100)
```

**b)** Below we perform a one sample t-test for the artificial data 
(that we generate ourselves).
```{r}
mu=0.2
x=rnorm(100,mu,1) # creating artificial data
t.test(x,mean=0)   # t.test(x,alternative=c("two.sided"),conf.level=0.95,mu=10)
```

**c)** We often do not need to report the whole output of R-commands, only certain values of the output. For example, below we perform a two-sample t-test and report only the (appropriately rounded) values of t-statistics and the p-pavue.

```{r}
mu=0;nu=0.5
x=rnorm(50,mu,1); y=rnorm(50,nu,1) # creating artificial data
ttest=t.test(x,y) 
```

The value of t-statistics in the above evaluation is  `r round(ttest[[1]],2)` and the p-value is `r round(ttest[[3]],4)`.
  
